---
title: "ORFik_Ribo-seq_pipeline"
author: "Haakon Tjeldnes & Kornel Labun"
date: "`r BiocStyle::doc_date()`"
package: "`r pkg_ver('ORFik')`"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{ORFik_Ribo-seq_pipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Welcome to the `ORFik` package. 
`ORFik` is an R package for analysis of transcript and translation features through manipulation of sequence data and NGS data.
This vignette will preview a simple Ribo-seq pipeline using ORFik. It is important you read all the other vignettes before this one, since functions will not be explained here in detail. 

# Pipeline
This pipeline will shows steps needed to analyse Ribo-seq from Alexaki et al, 2020
The following steps are done:

1. Define directory paths
2. Download Ribo-seq data from SRA (subset to 2 million reads)
3. Download annotation and contaminants
4. Align data
5. Make ORFik experiment
6. QC
7. Heatmaps

```{r eval = FALSE, echo = TRUE, message = FALSE}
#¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤#
# Ribo-seq HEK293 (2020) Investigative analysis of quality of new Ribo-seq data
#¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤#
# Article: https://f1000research.com/articles/9-174/v2#ref-5
library(ORFik)

#¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤#
# Settings (This is the only Custom part per user, rest you can just run)
#¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤#
# This is where you want your annotation and STAR index
annotation <- "/media/roler/S/data/Bio_data/references/Homo_sapiens_GRCh38_101/"
# Where to download fastq files
fastq.dir <- "/media/roler/S/data/Bio_data/raw_data/Ribo-seq/Alexaki_Human/"
# Where you want mapped bam files
bam.dir <- "/media/roler/S/data/Bio_data/processed_data/Ribo-seq/Alexaki_Human/"


#¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤#
# Download fastq files for experiment and rename (Skip if you have the files already)
#¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤#
# SRA Meta data download (work for ERA and DRA too)
study <- download.SRA.metadata("PRJNA591214", fastq.dir)
# Remove RNA-seq and CO libraries, we don't want those
study <- study[grep("mRNA-Seq|CO", sample_title, invert = TRUE),]
# Download fastq files (uses SRR numbers (Run column) from study)
download.SRA(study, fastq.dir, rename = FALSE, subset = 2000000)

# Which organism is this, scientific name, like "Homo sapiens" or "Danio rerio"
organism <- study$ScientificName[1] # Usually you find organism here, else set it yourself
paired.end <- study$LibraryLayout == "PAIRED"
paired.end.all <- all(paired.end)
#¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤#
# Annotation
#¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤#

annotation <- getGenomeAndAnnotation(
  organism = organism,
  genome = TRUE, GTF = TRUE,
  phix = TRUE, ncRNA = TRUE, tRNA = TRUE, rRNA = TRUE,
  output.dir = annotation,
  assembly_type = "primary_assembly"
)

#¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤#
# STAR index
#¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤#
# Remove max.ram = 20 and SAsparse = 2, if you have more than 64GB ram
index <- STAR.index(annotation, wait = TRUE, max.ram = 20, SAsparse = 2)

#¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤#
# Alignment (with depletion of phix, rRNA, ncRNA and tRNAs) & (with MultiQC of final STAR alignment)
#¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤#
alignment <-
  STAR.align.folder(fastq.dir, bam.dir, index,
                    paired.end = paired.end.all,
                    steps = "tr-co-ge", # (trim needed: adapters found, then genome)
                    adapter.sequence = "auto", # Adapters are auto detected
                    trim.front = 0, min.length = 20)

#¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤#
# Create experiment (Starting point if alignment is finished)
#¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤#
library(ORFik)
create.experiment(paste0(alignment, "/aligned/"), exper = "Alexaki_RFP",
                  fa = annotation["genome"],
                  txdb = paste0(annotation["gtf"], ".db"),
                  organism = organism,
                  pairedEndBam = paired.end,
                  viewTemplate = FALSE)
# Now fix experiment table in libre office
df <- read.experiment("Alexaki_RFP")

#¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤#
# Convert files and run Annotation vs alignment QC
#¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤#
ORFikQC(df)
#remove.experiments(df)
#¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤#
# P-shifting of Ribo-seq reads:
#¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤#
# From ORFikQC it looks like 20, 21, 27 and 28 are candidates for Ribosomal footprints
shiftFootprintsByExperiment(df, accepted.lengths = c(20:21, 27:28))



#¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤#
# Create heatmaps
#¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤#
# Remove RNA-seq
# Pre-pshifting
heatMapRegion(df, region = c("TIS", "TTS"), shifting = "5prime", type = "ofst",
              outdir = paste0(dirname(df$filepath[1]), "/QC_STATS/heatmaps/pre-pshift/"))
remove.experiments(df)
# After pshifting
heatMapRegion(df, region = c("TIS", "TTS"), shifting = "5prime", type = "pshifted",
              outdir = paste0(dirname(df$filepath[1]), "/QC_STATS/heatmaps/pshifted/"))
```
