---
title: "Working with transcripts"
author: "Haakon Tjeldnes & Kornel Labun"
date: "`r BiocStyle::doc_date()`"
package: "`r pkg_ver('ORFik')`"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Working with transcripts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction
Welcome to the introduction to working with transcripts in ORFik. This vignette will walk you through how
to find and manipulate the sections of transcript you want. And how to calculate coverage over these regions.
`ORFik` is an R package containing various functions for analysis of RiboSeq, RNASeq, RCP-seq, TCP-seq, Chip-seq and Cage data, we advice you to read Importing data vignette, before starting this one.

## Motivation
Working with all the specific regions of transcripts can be hard. The standard Bioconductor
packages GenomicRanges and GenomicFeatures does not solve most of these problems, so
ORFik extends this ecosystem with what we call TranscriptRanges and TranscriptFeatures.
We will here learn how to use ORFik
effectively getting the transcript regions we want and compute coverage over those regions etc.

# Getting data
Let us use the same data we used in the importing data vignette.

```{r eval = TRUE, echo = TRUE, message = FALSE}
library(ORFik)
organism <- "Saccharomyces cerevisiae" # Baker's yeast
output_dir <- tempdir() # Let's just store it to temp
# If you already downloaded, it will not redownload, but reuse those files.
annotation <- getGenomeAndAnnotation(
                    organism = organism,
                    output.dir = output_dir,
                    assembly_type = "toplevel"
                    )
genome <- annotation["genome"]
gtf <- annotation["gtf"]
txdb_path <- paste0(gtf, ".db")
txdb <- loadTxdb(txdb_path)
```

# Loading Transcript region (single point locations)

Let us get these regions:

- Transcription start sites (TSS, start of leaders)
- Translation start sites (TIS, start of CDSs)
- Translation termination sites (TTS, end of CDSs)
- Transcription end sites (TES, end of mRNA)
```{r eval = TRUE, echo = TRUE, message = FALSE}
## TSS
startSites(loadRegion(txdb, "leaders"), asGR = TRUE, is.sorted = TRUE)
# Equal to:
startSites(loadRegion(txdb, "mrna"), asGR = TRUE, is.sorted = TRUE)
## TIS
startSites(loadRegion(txdb, "cds"), asGR = TRUE, is.sorted = TRUE)
## TTS
stopSites(loadRegion(txdb, "cds"), asGR = TRUE, is.sorted = TRUE)
## TES
stopSites(loadRegion(txdb, "mrna"), asGR = TRUE, is.sorted = TRUE)
```
# Loading Transcript region (window locations)
What if we want a window around that region point?
Then we also need might the mrna region information,
to map to transcript coordinates and get area around that point
```{r eval = TRUE, echo = TRUE, message = FALSE}
mrna <- loadRegion(txdb, "mrna")
## TSS
startRegion(loadRegion(txdb, "mrna"), upstream = 30, downstream = 30, is.sorted = TRUE)
## TIS
startRegion(loadRegion(txdb, "cds"), mrna, upstream = 30, downstream = 30, is.sorted = TRUE)
## TTS
stopRegion(loadRegion(txdb, "cds"), mrna, upstream = 30, downstream = 30, is.sorted = TRUE)
## TES
stopRegion(loadRegion(txdb, "mrna"), upstream = 30, downstream = 30, is.sorted = TRUE)
```
# Extending transcript into genomic flanks
You might notice that since we are doing transcript coordinates, it will not actually extend 30 upstream
from TSS, because there is no transcript there.

To extend into the genomic region do:
```{r eval = TRUE, echo = TRUE, message = FALSE}
# TSS genomic extension
extendLeaders(startRegion(mrna[1], upstream = -1, downstream = 30, is.sorted = TRUE), 30)
# TES genomic extension
extendTrailers(stopRegion(mrna[1], upstream = 30, downstream = 0, is.sorted = TRUE), 30)
```

# Coverage over transcript regions

So how could we now get the coverage per nt in the TIS region of a cds?

## Loading NGS data
We first need some NGS data, so let us now use the ORFik experiment included in the package:
```{r eval = TRUE, echo = TRUE, message = FALSE}
df <- ORFik.template.experiment()
df
```
There are 4 CAGE libraries, 4 PAS (Poly-A seq) libraries, 4 Ribo-seq libraries and
4 RNA-seq libraries.
Let's load the first Ribo-seq library in this artificial study
```{r eval = TRUE, echo = TRUE, message = FALSE}
df <- df[9,]
df
```

Load the Ribo-seq
```{r eval = TRUE, echo = TRUE, message = FALSE}
ribo <- fimport(filepath(df, "default"))
```

## Total counts per region/window
```{r eval = TRUE, echo = TRUE, message = FALSE}
TIS_window <- startRegion(loadRegion(df, "cds"), loadRegion(df, "mrna"), 
                is.sorted = TRUE, upstream = 20, downstream = 20)
```


Now let's get the total counts per window
```{r eval = TRUE, echo = TRUE, message = FALSE}

countOverlapsW(TIS_window, ribo, weight = "score")
# This is shorter version (You do not need TIS_window defined first) ->
startRegionCoverage(loadRegion(df, "cds"), ribo, loadRegion(df, "mrna"), 
                    is.sorted = TRUE, upstream = 20, downstream = 20)
```
You see the first gene had 33 reads in the 20-20 window around TIS

## Per nucleotide in region/window
But what if you want the coverage per nucleotide in that window?

```{r eval = TRUE, echo = TRUE, message = FALSE}
# TIS region coverage
coveragePerTiling(TIS_window, ribo)[1:3]
```
For plotting it is best to use the data.table return:

```{r eval = TRUE, echo = TRUE, message = FALSE}
# TIS region coverage
dt <- coveragePerTiling(TIS_window, ribo, as.data.table = TRUE)
head(dt)
```
Let's plot the coverage for fun, now we want frame information too:

```{r eval = TRUE, echo = TRUE, message = FALSE}
# TIS region coverage
dt <- coveragePerTiling(TIS_window, ribo, as.data.table = TRUE, withFrames = TRUE)
# Rescale 0 to position 21
dt[, position := position - 21]
pSitePlot(dt)
```
